Reducing Acceptance Marks \\ in Emerson-Lei Automata

### 2021-10-14 paper/main.tex
Intro

%{{{ Abstract

\begin{abstract}
  The formalism called \emph{transition-based Emerson-Lei automata
    (TELA)} generalizes many traditional kinds of automata over
  infinite words including Büchi, co-Büchi, Rabin, Streett, and parity
  automata. Transitions in TELA are marked with acceptance marks and
  accepting formulas are boolean combinations of terms saying that a
  particular mark has to be visited infinitely or finitely often in an
  accepting run. Algorithms processing these automata can be very
  sensitive to the number of different acceptance marks. We introduce
  two techniques reducing the number of acceptance marks in a TELA
  without altering its structure and the represented language. One
  technique represents a standard approach based on relations among
  acceptance marks in individual strongly connected components of the
  automaton. The other technique constructs quantified boolean formula
  (QBF) queries that ask a QBF solver for an acceptance condition with
  fewer acceptance marks and the placement of these marks in the
  automaton structure. Both techniques are implemented and experiments
  show that the number of acceptance marks in TELA produced by
  state-of-the-art tools ltl3tela, Rabinizer~4, and SPOT can be
  sometimes reduced. In the case of Rabinizer~4, we reduced the
  cummulative number of acceptance marks to less than one half.
\end{abstract}

@@

\section{Introduction}
\begin{itemize}
\item Emerson-Lei automata jsou v principu staré, ale v poslední době
  znovuobjevene~\cite{babiak.15.cav} a je o tom ted spousta
  clanku~\cite{baier.19.atva} a nastroju
\item algoritmy jsou obvykle citlive na akceptacni podminku (citace)
\item cilem prace je zjednodusit akceptacni podminku (zejmnena
  redukovat pocet akceptacnich znacek) bez zmeny struktury automatu
\end{itemize}

\section{Preliminaries}

\subsection{TELA}
A \emph{transition-based Emerson-Lei automaton} over alphabet $\Sigma$ is a tuple $\mathcal{A} = \left(  \mathcal{Q}, M, \Sigma,q_{0}, \delta, \varphi \right)$, where
\begin{itemize}

    \item $\mathcal{Q}$ is a finite set of \emph{states},
    \item $M$ is a finite set of \emph{acceptance marks},
    \item $\Sigma$ is a finite \emph{alphabet},
    \item $q_{0}$ is an \emph{initial state},
    \item $\delta \subseteq \mathcal{Q} \times \Sigma \times 2^{M} \times \mathcal{Q}$ is a transition relation,
    \item $\varphi$ is the acceptance condition constructed by the following abstract syntax equation, where $m \in M$:

    \begin{center}
        $\varphi ::= \true \mid \false \mid \Inf(m) \mid \Fin(m) \mid \left( \varphi \land \varphi \right) \mid \left( \varphi \lor \varphi \right)$.\\
    \end{center} 
\end{itemize}

@@

\section{Reduction based on acceptance sets relations / SCC-based reduction}
In this section, the simplification strategy is to find relations between acceptance sets and reduce the redundant ones. SCC-based simplification is the coarsest simplification we propose.

@@

\section{QBF-based reduction}
\begin{itemize}
\item obecny popis konstrukce formule
\item (level 3)
\item level 4
\end{itemize}

\section{Implementation and experimental evaluation}

\section{Conclusions}

>>>>

%{{{ abstract

\begin{abstract}
  The formalism called \emph{transition-based Emerson-Lei automata
    (TELA)} generalizes many traditional kinds of automata over
  infinite words including Büchi, co-Büchi, Rabin, Streett, and parity
  automata. Transitions in TELA are marked with acceptance marks and
  accepting formulas are boolean combinations of terms saying that a
  particular mark has to be visited infinitely or finitely often in an
  accepting run. Algorithms processing these automata can be very
  sensitive to the number of different acceptance marks. We introduce
  two techniques reducing the number of acceptance marks in a TELA
  without altering its structure and the represented language. One
  technique represents a standard approach based on relations among
  acceptance marks in individual strongly connected components of the
  automaton. The other technique constructs \emph{quantified boolean
    formula (QBF)} queries that ask a QBF solver for an acceptance
  condition with fewer acceptance marks and the placement of these
  marks in the automaton structure. Both techniques are implemented
  and experiments show that the number of acceptance marks in TELA
  produced by state-of-the-art tools ltl3tela, Rabinizer~4, and SPOT
  can be sometimes reduced. In the case of Rabinizer~4, we reduced the
  cummulative number of acceptance marks to less than one half.
\end{abstract}

@@

%{{{ introduction

\section{Introduction}

% Automata over infinite words (also known as $\omega$-automata) are now
% used in specification, verification, analysis, monitoring, and
% synthesis of various systems with infinite behaviour. The first and
% the most common kind of $\omega$-automata is Büchi
% automaton~\cite{buchi.60.clmps} introduced in 1960. As deterministic
% Büchi automata are less expressive than nondeterministic ones,
% deterministic $\omega$-automata are typically considered with other
% acceptance condition like Rabin, Streett, Muller, or parity. In 1987,
% Emerson and Lei~\cite{emerson.87.scp} introduced $\omega$-automata
% where the acceptance condition can be an arbitrary combination of
% acceptance primitives saying that a certain set of states should be
% visited finitely often or infinitely often. In 2015, the same kind of
% acceptance condition was implemented in the \emph{Hanoi omega-automata
%   format (HOAF)}~\cite{babiak.15.cav}, with the difference that finiteness

In 1987, Emerson and Lei~\cite{emerson.87.scp} introduced automata
over infinite words where acceptance conditions are arbitrary
combinations of acceptance primitives saying that a certain set of
states should be visited finitely often or infinitely often. In 2015,
the same kind of acceptance condition was described in the \emph{Hanoi
  omega-automata format (HOAF)}~\cite{babiak.15.cav}. The only
difference is that the acceptance primitives talk about finitely or
infinitely often visited acceptance marks rather than sets of states.
Acceptance marks are placed on transitions and each mark identifies
the set of transitions containing this mark. Hence, these automata are
called \emph{transition-based Emerson-Lei automata (TELA)} and they
generalize many traditional kinds of automata over infinite words
including Büchi, co-Büchi, Rabin, Streett, Muller or parity
automata.

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use less states than equivalent automata with traditional
acceptance conditions. Further, algorithms handling TELA can
automatically handle all automata with traditional acceptance
conditions. TELA automata can be obtained for example by translating
LTL formulas by Delag~\cite{muller.17.gandalf},
ltl3tela~\cite{major.19.atva}, Rabinizer~4~\cite{kretinsky.18.cav}, or
SPOT~\cite{duret.16.atva2}. There are also some algorithms processing
these automata, for example the emptiness check~\cite{baier.19.atva}
or translation of TELA to parity automata~\cite{renkin.20.atva}.

\todo{Nechcem do experimentu pridat Delag?}

Algorithms processing TELA are often sensitive to the number of
acceptance marks more than to other parts of the automaton. For
example, the transformation of TELA to parity automata based on
\emph{color appearance record}~\cite{renkin.20.atva} transforms a TELA
with $s$ states and $m$ acceptance marks into a parity automaton with
up to $s\cdot m!$ states. Further, the emptiness
check~\cite{baier.19.atva} is exponential in the number of acceptance
marks that appear in acceptance primitives saying that a mark has to
be visited finitely often, while it is only polynomial in other
measures of the input automaton.

These observations motivate our study of techniques reducing the
number of acceptance marks without altering the structure of the
automaton. We present basically two reduction techniques. The first
one is based on relations among individual transition sets defined by
various acceptance marks in each \emph{strongly connected component
  (SCC)}. This technique can be seen as an extension of the
\emph{SCC-base simplifications of acceptance
  conditions}~\cite{babiak.13.spin} designed for
\emph{transition-based generalized Büchi automata (TGBA)} and
implemented in SPOT. Probably the main difference between the original
technique for TGBA and the one for TELA is in the final step when
simplified acceptance conditions for each SCC are merged together into
one reduced acceptance condition.

The second technique is completely different. For a given TELA and
parameters $k,n$, it produces a \emph{quantified boolean formula
  (QBF)} which is satisfiable if and only if there exists an
equivalent automaton with the identical structure, $n$ acceptance
marks, and an acceptance formula in \todo{conjunctive?}  normal form
with $k$ clauses. The placement of the marks on transitions and the
acceptance formula can be obtained from a model of the
formula. Besides this formula, we describe also the construction of a
simpler formula such its satisfiability implies existence of an
equivalent automaton with the identical structure and $n$ acceptacne
marks, but not vice versa.

We have implemented both reduction techniques in a tool called
\texttt{telatko}. We show that the tool can reduce acceptance marks in
automata produced by ltl3tela~\cite{major.19.atva},
Rabinizer~4~\cite{kretinsky.18.cav}, and SPOT~\cite{duret.16.atva2}.
While the reduction is relatively modest for TELA produced by ltl3tela
and SPOT, it is substantial for automata produced by Rabinizer~4.

\subsubsection{Related results}
We are not aware about any existing research focused on simplification
of acceptance formulas of TELA or reduction of the number of its
acceptance marks without increasing the number of states.

Casares, Colcombet, and Fijalkow very recently introduced a structure
called \emph{alternating cycle decomposition
  (ACD)}~\cite{casares.21.icalp} which compactly represents the
information about all accepting and non-accepting automata cycles.  We
expect that ACD could potentialy be used to reduce the number of
acceptance marks or to simplify the acceptance condition. However,
such a reduction is not obvious. Moreover, the construction of ACD
seems to be computationally expensive.

\subsubsection{Structure of the paper} The next section introduces
basic terms used in the paper. Section~\ref{sec:standard} presents the
reduction technique based on relations among individual acceptance
marks on each SCC. The QBF-based technique is explained in
Section~\ref{sec:qbf}. Implementation is briefly described in
Section~\ref{sec:experiments}, which also shows the experimental
results. Finally, Section~\ref{sec:conclusion} suggests other
applications of our QBF-reduction technique and closes the paper.

%}}}
%{{{ preliminaries

\section{Preliminaries}\label{sec:prelim}

In this section we recall the basic terms around TELA and QBF.

\begin{definition}[TELA]
  A \emph{transition-based Emerson-Lei automaton (TELA)} is a tuple
  $\mathcal{A} =(Q,M,\Sigma,q_{0},\delta,\varphi)$, where
  \begin{itemize}
  \item $Q$ is a finite set of \emph{states},
  \item $M$ is a finite set of \emph{acceptance marks},
  \item $\Sigma$ is a finite \emph{alphabet},
  \item $q_{0}\in Q$ is an \emph{initial state},
  \item $\delta \subseteq Q \times \Sigma \times 2^{M} \times Q$ is a
    \emph{transition relation}, and
  \item $\varphi$ is the \emph{acceptance condition} constructed by
    the following abstract syntax equation, where $m$ ranges over $M$:

    \begin{center}
        $\varphi ::= \true \mid \false \mid \Inf(m) \mid \Fin(m) \mid \left( \varphi \land \varphi \right) \mid \left( \varphi \lor \varphi \right)$.\\
    \end{center} 
\end{itemize}
\end{definition}

\todo{Honza je tu a jde spat (je 5 rano pryc).}

@@

%}}}
%{{{ standard reductions

\section{Reduction based on acceptance marks relations}\label{sec:standard}

In this section, the simplification strategy is to find relations between acceptance sets and reduce the redundant ones. SCC-based simplification is the coarsest simplification we propose.

@@

%}}}
%{{{ QBF-based reduction

\section{QBF-based reduction}\label{sec:qbf}

\begin{itemize}
\item obecny popis konstrukce formule
\item (level 3)
\item level 4
\end{itemize}

%}}}
%{{{ implementation and experimental evaluation

\section{Implementation and experimental evaluation}\label{sec:experiments}

%}}}
%{{{ conclusion

\section{Conclusions}\label{sec:conclusion}
